---
title: "CERMBweather package introduction"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)

dbpath <- "c:/michael/Rworkspaces/BOMdata/database/test.db"
if (file.exists(dbpath)) unlink(dbpath)

```

## Installing the package

The current version of the package can be installed from GitHub with the following R commands:

```{r eval = FALSE}

# If you don't already have the devtools package
install.packages("devtools")

devtools::install_github("mbedward/CERMBweather")

```

You will need the following packages installed:

  * DBI - general database functions
  * dbplyr - allows dplyr to be used with database tables
  * dplyr - functions to work with data frames
  * RSQLite - specific functions for SQLite databases
  * stringr - functions to query and manipulate character strings
  
These should be included as part of the install from GitHub but if not, the following command will install the current version of each:

```{r eval = FALSE}

install.packages(c("DBI", "dbplyr", "dplyr", "RSQLite", "stringr"))

```


## Functions to work with BOM zip files

There are several functions to work with zip files of synoptic or AWS (ie. hourly or half-hourly) weather data as received from BOM. The zip files contain CSV format data files, one per weather station.

### Summary of zip file contents

```{r}

library(CERMBweather)
library(dplyr, warn.conflicts = FALSE)

# Edit these paths for your system
#DIR <- "s:/Biosci Bushfires/CERMB_LIBRARY/DATA/WEATHER/BOM_updates/"
DIR <- "c:/michael/Rworkspaces/BOMdata"

SYNOPTIC <- file.path(DIR, "updates", "DS106_Clarke_synoptic_20180521.zip")
AWS <- file.path(DIR, "updates", "DS106_Clarke_halfhourly_20180521.zip")


# list the weather stations with non-empty data sets and 
# display the first few rows
stns <- bom_zip_summary(SYNOPTIC)
head(stns)

```

```{r}

# Find any weather stations in the zip file that have
# empty data sets
bom_zip_summary(SYNOPTIC, include = "empty")

```

### Retrieve station data directly

The function `bom_station_data` extracts data for one or more weather stations from a zip file and makes it available directly in your R workspace. By default, the function returns a list of data frames. Each data frame contains a subset of the BOM raw data fields (e.g. 'quality' fields are omitted) with more concise column names. 


```{r}

# Load data for three specified stations
res <- bom_zip_data(SYNOPTIC, stations = c(3080, 14840, 200790))

# List element(s) are named by station number, formatted as a
# six charcter string
names(res)

```

We can access a single data frame with the usual R list operations. Any of the following will retrieve the second data set (for station 14840):

```{r}

dat <- res[[2]]
dat <- res[["014840"]]
dat <- res$`014840`

```

You can also use the optional `out.format` argument to return data for multiple stations as a single data frame.

```{r}

dat <- bom_zip_data(SYNOPTIC, stations = c(3080, 14840, 200790), 
                    out.format = "single")

dim(dat)

# number of records for each stations
table(dat$station)

# first few records
head(dat)

```


## Database functions

Most of the time you will want to transfer weather station data from zip files, directories or individual files into a database (e.g. a central copy on the share drive) rather than load them directly into your R workspace. The package uses [SQLite](https://www.sqlite.org) as the database back-end. All data are stored in a single file which can be moved or copied as desired. SQLite databases have full support for SQL and can be accessed from a wide range of applications. 

In R, tables in a SQLite database can be queried in much the same way as when working with data frames using the `dplyr` and `dbplyr` packages. They can also be queried from R using SQL commands with the `DBI` package. The `CERMBweather` package provides functions that hide the messy detail for the most common tasks.


### Creating a database

The code below creates a new database with a specified path and file name (SQLite does not require any particular file name convention) containing tables for synoptic and AWS data.

```{r}

# Create a new database with the required tables
con <- bom_db_init("c:/michael/Rworkspaces/BOMdata/database/test.db")

```

The object returned by the function is a database connection (an object of class `SQLiteConnection`). We can use it to import data into the database and, later, to query records in the database.


### Importing data from zip files

In the code below, `SYNOPTIC` and `AWS` are the file paths defined earlier for BOM zip files.

Import data from the synoptic zip file:

```{r}

bom_db_import(con, SYNOPTIC)

```


Import data from the AWS zip file:

```{r}

bom_db_import(con, AWS)

```

The import function silently ignores any weather station records that are already present in the database, so if we repeat the last import we should see 0 records imported:

```{r}

bom_db_import(con, AWS)

```


### Importing data from files in a directory

When passed a directory path rather than a zip file path, the `bom_db_import` function will import data from all, or a specified subset, of the weather station data files in that directory.

```{r}

path <- file.path(DIR, "rawfiles", "Synoptic")

bom_db_import(con, path)

```


### Importing data from an individual file

The `bom_db_import` will also work with an individual file.

```{r}

path <- file.path(DIR, "rawfiles", "AWS", "HM01X_Data_200284_999999999515424.txt")

bom_db_import(con, path)

```



### Querying the database

Once we have some data in the database we can extract or summarize data in a variety of ways. The package provides functions to summarize the contents of the database.

Count synoptic and AWS records:

```{r}

bom_db_summary(con)

```

Count records by station:

```{r}

dat <- bom_db_summary(con, by = "station")

head(dat, 3)
tail(dat, 3)

```

Arbitrary queries can be done using `dplyr` package functions and `tbl` objects which act as proxies for database tables, allowing them to be treated like data frames.

The code below finds the maximum recorded temperature for each weather station.

```{r}

# Get a dplyr `tbl` object for the Synoptic table
tsynoptic <- bom_db_tbl(con, "syn")

# Find maximum temperature for each station
dat <- tsynoptic %>%
  # construct the query
  group_by(station) %>%
  summarize(maxtemp = max(temperature, na.rm = TRUE)) %>%
  filter(!is.na(maxtemp)) %>%
  
  # execute the query
  collect()

# Display results
plot(density(dat$maxtemp), 
     xlab = "temperature", ylab = "density",
     main = "Maximum recorded temperatute at each station")

```


The query above could also be performed using an SQL statement via the `DBI` package.

```{r}

qry <- paste(
  "SELECT station, MAX(temperature) AS maxtemp FROM SYNOPTIC",
  "GROUP BY STATION",
  "HAVING NOT(maxtemp IS NULL)")

dat2 <- DBI::dbGetQuery(con, qry)

```

This query produces the same result as the earlier dplyr query. 

```{r}

all.equal(dat, dat2)

```



### Closing the database connection

Closing a connection will also close any `tbl` objects associated with it.

```{r}

bom_db_close(con)

```

